---
title: "main"
output: html_document
---


```{r load packages}
#---------i. Clear environment
rm(list=ls())
#---------ii.Load here package to enable finding script loading other package
# MISSING Add Package loader from forecasting prokect
# Maybe using a pip freeze type thing (packagerat?)


```

```{r}
library(here)
source(here("R", "load_snb_data.R"))
source(here("R", "utils.R"))
```

```{r}
source(here("scripts", "01_load_data.R"))
```

# Clean Data and organzie into one dataframe

General questions
  How do I handle dates? Thinking using ym since some data is mothly, handling the missing values with Kalman filter
  But is that good for the compute? Or even reasonable


## Clean SNB Money market data
```{r}
library(tidyverse)

mm_raw <- mm_data[[1]]

mm_raw <- mm_data %>%
  pluck("data") %>%
  bind_rows()

mm_raw_wide <- mm_raw %>%
  pivot_wider(
    names_from = "D0",
    values_from = "Value"
  )

saron_ts <- standardize_time_series_df(mm_raw_wide, "Date", "SARON", "saron", "%Y-%m")
libor_ts <- standardize_time_series_df(mm_raw_wide, "Date", "3M0", "3m_libor", "%Y-%m")

```

  

# Clean SNB Gov Bond Data
```{r}

gb_raw <- gb_data[[1]]

gb_raw <- gb_data %>%
  pluck("data") %>%
  bind_rows()

gb_raw_wide <- gb_raw %>%
  pivot_wider(
    names_from = "D0",
    values_from = "Value"
)


gb_5y_ts <- standardize_time_series_df(gb_raw_wide, "Date", "5J", "5y_bond", "%Y-%m")
gb_10y_ts <- standardize_time_series_df(gb_raw_wide, "Date", "10J", "10y_bond", "%Y-%m")

```

Access data from list
  select date and values
  rename columns to date and gov_bond_rate_{duration}y
  turn dates into yq
  turn rates into floats(assurance)
  
  
# Clean REER

First need to find a way to download it



# Clean CPI Data
```{r}

```

Decide which series to take (what date is 100)

  select date and values
  rename columns to date and money_market_rate
  turn dates into ym (can always be turned into yq later)
  turn rates into floats(assurance)
  

# Clean Unemployment Data
```{r}

```


select region total
select date column annd values
rename columns to date and unemp_rate
turn date into ym,
turn values in floats


# Clean employment Data
```{r}

```

Select 5-96 code (excludes agriculture) and then drop column
select sex total and then drop column
rename columns to date and emp_tot
select dates and values
turn date into yq
turn values into floast 


# clean GDP Data
```{r}

```

Values already selected in download -> move here and do that in cleaning
still need to:
  drop unnecessary columns: structure, type, seas_adj
  rename columns to date and gdp
  convert date into yq
  convert values in floats






