---
title: "SSM_Documentation"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library("here") # if error, install it
source(here("R", "load_snb_data.R"))
source(here("R", "load_kof_data.R"))
source(here("R", "utils.R"))
source(here("R", "kalman_builder.R"))
source(here("R", "kalman_procedures.R"))
source(here("R", "ssm_forecasting.R"))

```

# State-Space Model (SSM) Documentation

This document outlines the theoretical framework and implementation
details of the State-Space Model and the Kalman Filter used in this
project.

## 1. Model Specification

The system is defined by two fundamental equations that describe the
relationship between observed data and hidden structural components.

### Measurement Equation

The measurement equation relates the observed variables $y_t$ to the
latent state $\rho_t$:

$$y_t = \mu_t + G\rho_t + M\varepsilon_t \tag{1}$$

-   $\mu_t$: A vector representing the intercept or the contribution of
    exogenous variables (e.g., the Okun’s Law cyclical impact $X\beta$).
-   $G$: The factor loading matrix that maps the latent state to the
    observed measurements.
-   $M$: The coefficient matrix for the measurement noise
    $\varepsilon_t$, defining the observation error variance-covariance
    structure.

### Transition Equation

The transition (or state) equation describes the evolution of the latent
state over time:

$$\rho_t = \nu_t + H\rho_{t-1} + N\xi_t \tag{2}$$

-   $\nu_t$: The intercept of the state equation, typically used to
    represent a deterministic drift in the trend.
-   $H$: The transition matrix defining the persistence of the state
    from $t-1$ to $t$.
-   $N$: The coefficient matrix for the state innovation $\xi_t$,
    defining the process noise variance.

## 2. Estimation Pipeline

The core estimation logic is contained within the `kalman_filter()`
function located in `kalman_procedures.R`. This function serves as the
engine, processing system matrices to generate filtered state estimates
and the corresponding log-likelihood for optimization.

### Matrix Construction

System matrices are constructed externally via "factory" functions
(e.g., `mu_t_matrix_factory()`, `G_matrix_factory()`). These matrices
output a builder function (e.g., `mu_t_builder()`, `G_builder())`, that
builds the final matrices with the parameters from the optimizer. This
modular approach allows injection of time-varying exogenous data and
specific parameter constraints into the matrices before they are passed
to the filter. This ensures that the filter remains a pure mathematical
operator, independent of the specific model configuration.

### Filter Inputs

Beyond the primary system matrices ($G, H, M, N, \mu_t, \nu_t$), the
estimation pipeline handles several advanced configurations:

-   **Initial Conditions**: Requires initial guesses for the state
    $\rho_0$ and the state covariance $\Sigma_0$.

-   **Constraints**: Supports `indic_pos` to enforce positivity
    constraints on specific parameters during optimization.

-   **Reconciliation**: Allows for functional adjustments to the state
    through reconciliation functions. (Currently not needed)

-   **Quadratic Filtering**: The architecture supports $Q$ functions for
    quadratic filter extensions. These are currently not needed

The pipeline operates by iteratively updating the state estimate as new
data arrives, minimizing the prediction error and maximizing the
log-likelihood of the observed series $Y$.

```{r}
# uncomment to see the function
# kalman_filter
```

# Optimization Strategy

The parameter estimation for the State-Space Model is conducted using a
maximum likelihood approach, implemented via numeric optimization.

## 1. The Objective Function

The optimization process utilizes the `optimx` package to minimize the
negative log-likelihood (or maximize the log-likelihood) of the system.
The objective function, `loglik_ssm`, acts as a bridge between the
optimizer and the Kalman Filter:

-   **Matrix Mapping**: It receives a vector of candidate parameters
    from the optimizer and maps them into the structured system matrices
    ($G, H, M, N, \mu_t, \nu_t$) via their builder functions.

-   **Filtering**: These matrices are passed into the `kalman_filter`
    function.

-   **Likelihood Calculation**: The filter processes the observed data
    $Y$ and returns the log-likelihood based on the prediction errors
    (innovations) and their associated variances.

## 2. Recursive Optimization Flow

The pipeline follows an iterative logic to converge on the optimal
parameter set:

-   **Initialization**: The process begins with either a user-defined
    initial guess or the results of a previous estimation.

-   **Evaluation**: The Kalman Filter computes the likelihood of the
    data given the current parameter guess.

-   **Adjustment**: The optimizer analyzes the gradient and curvature of
    the likelihood surface to propose a new, improved parameter vector.

-   **Convergence**: The loop continues until the marginal improvement
    in the log-likelihood falls below a specified tolerance threshold,
    at which point the final estimates are returned.

## 3. The Estimation Wrapper

To ensure robust convergence and prevent the model from becoming trapped
in local optima, a dedicated wrapper function manages the optimization
loops.

### Multi-Method Sequential Optimization

The wrapper allows for the execution of multiple optimization algorithms
(e.g., Nelder-Mead followed by BFGS) in a sequential chain. In this
setup, the final parameter output of the first method serves as the
starting value for the second. This hybrid approach allows us to use
different methods with different strengths.

### Iterative Refinement

The user can specify multiple global loops. Each iteration begins with
the best parameters found in the previous cycle, progressively refining
the estimation. This iterative "tidying" of the parameters ensures that
the final results represent a stable, globally optimized configuration
for the State-Space Model.

```{r}
loglik_ssm
```

```{r}
ssm_optimizer_wrapper
```

# Matrix Handling and Function Factories

The primary technical challenge in the estimation pipeline is the
dynamic nature of the system matrices. Since these matrices must be
updated with new candidate parameters at every iteration of the
optimizer, they cannot be static objects. This is resolved through the
use of **Function Factories**.

A function factory is a higher-order function that generates a specific
"builder" function. This builder is pre-configured with the model's
structural specifications—such as the number of exogenous variables or
the type of stochastic process (e.g., Random Walk vs. AR(1)). Once
created, the builder function accepts a vector of parameters from the
optimizer and populates the final matrices required by the Kalman
Filter.

## 1. Building $\mu_t$ (Exogenous Effects)

The $\mu_t$ matrix is the most complex component to construct as it
incorporates time-varying exogenous data. It functions similarly to an
intercept in a regression model but can vary at each time step $t$.

The builder generates a matrix where each row corresponds to an
observation in time $T = 1, \dots, T_{end}$, and each column corresponds
to a measurement variable in $Y$. During the filtering process, the
algorithm selects the appropriate row for the current time step. For a
model with two measurement variables and several exogenous predictors
(such as Okun’s Law features), the matrix is structured as:

$$
  \mu_t = 
  \begin{bmatrix}
    \beta_0 + \beta_1 X_{1,1} + \beta_2 X_{1,2} & 0 \\
    \beta_0 + \beta_1 X_{2,1} + \beta_2 X_{2,2} & 0 \\
    \vdots & \vdots \\
    \beta_0 + \beta_1 X_{T,1} + \beta_2 X_{T,2} & 0
  \end{bmatrix}
$$ In context of the Okuns Law model, the first row is the first
observation and we see that the first column with the exogenous
regressors impacts the first measurement variable, which is
unemployment. The second measurement variable is not impacted by
anything so the second column is 0.

The factory also produces a **manifest**, a list of parameters to be
estimated, including initial "educated guesses" as starting points for
the optimizer and boundary constraints.

```{r}
mu_t_matrix_factory
```

## 2. Building $G$ (Factor Loadings)

Matrix $G$ governs the link between the latent state $\rho_t$ and the
measurement equation. Its dimensions are determined by the number of
states and the number of observed variables. In the current
specification, $G$ typically defaults to a vector of ones, implying a
1-to-1 relationship between the state (the natural rate) and the
observations. While currently static, the factory architecture allows
for future implementations where loadings can be estimated as
parameters.

```{r}
G_matrix_factory
```

## 3. Building $M$ (Measurement Noise)

Matrix $M$ represents the measurement error covariance, quantifying the
"noise" or uncertainty in each observed variable. We assume independent
noise, resulting in a diagonal matrix:

$$M = \begin{bmatrix}
\sigma_{\epsilon, 1} & 0 \\ 0 & \sigma_{\epsilon, 2}
\end{bmatrix}$$

To ensure that variances remain strictly positive ($0 < \sigma$), the
builder applies an exponential transformation, $\exp(\theta)$, to the
optimizer's candidate values. If the estimated variances are small, the
Kalman Filter places high "trust" in the observed data; conversely, high
variances cause the filter to rely more heavily on the state transition
logic.

```{r}
M_matrix_factory
```

## 4. Building $H$ (State Transition)

The $H$ matrix determines the persistence of the state $\rho$ from $t-1$
to $t$. The structure depends on the specified process: - **Random
Walk**: $H = [1]$. The state is structural, and changes are driven
exclusively by shocks ($\xi_t$). - **AR(1) Process**: $H = [\phi]$,
where the persistence parameter $0 < \phi < 1$ must be estimated.

```{r}
H_matrix_factory
```

The logic of the random walk for the specification in the model is that
the rate is structural and only shocks to the natural rate change it.
These shocks are then governed by the matrix N

## 5. Building $N$ (State Innovation)

The $N$ matrix governs the variance of the shocks to the state equation.
Currently, $N$ is specified manually within the `kalman_builder.R` or
`loglik_ssm()` functions. Future iterations will migrate this into a
dedicated factory function to allow the optimizer to estimate the
volatility of the natural rate directly.

# The Manifest Logic

Every factory function, in addition to producing a builder, generates a
**Manifest**. The manifest serves as a centralized metadata repository
that defines the search space for the numerical optimizer.

## 1. Parameter Metadata and Constraints

The manifest explicitly maps each estimated coefficient to its
corresponding role in the system matrices. It contains:

-   **Initial Guesses**: Pre-defined starting values (e.g., based on
    historical averages or economic theory) that provide the optimizer
    with a stable entry point into the likelihood surface.

-   **Boundary Constraints**: Specific logical limits for parameters,
    such as enforcing that the AR(1) coefficient $\phi$ remains between
    0 and 1, or that variance terms $\sigma$ are strictly positive.

## 2. Integration with the Optimizer

The manifest ensures that the optimizer remains "aware" of the model's
structural requirements. By centralizing these definitions, the manifest
allows for rapid model iteration; for instance, if an additional
exogenous variable is added to $\mu_t$, the manifest automatically
updates the number of parameters the optimizer needs to track. This
creates a self-documenting pipeline where the model's structural
assumptions are always aligned with the estimation process.

# Rolling Pseudo-Out-of-Sample Forecast Logic

The forecasting framework is implemented in two distinct phases:
**Rolling Parameter Estimation** and **Path Projection**.

## 1. Temporal Framework

We define the time dimension of our sample as $T \in [1, T_{end}]$. For
any given forecast origin $t$, the model parameters are estimated using
the information set $\mathcal{I}_t = \{y_1, \dots, y_t\}$. From this
vantage point (pseudo present where we forecast from), we project the
unemployment rate forward for a horizon $h = \{1, \dots, 8\}$,
corresponding to two years. Once the forecast at origin $t$ is complete,
the vantage point advances to $t+1$, where the model is re-estimated
using $\mathcal{I}_{t+1}$, and a new path is generated. This recursive
process repeats until $\mathcal{I} = T\_{end}$ reaches the end of the
available estimation sample.

## 2. Function Workflow

The logic is split between two primary functions. First,
`get_ssm_forecast_parameters()` executes the rolling estimation,
yielding a sequence of optimal parameter vectors $\theta_t$ and the
filtered state (natural rate) $\rho_t$ (here the natural rate
$\bar{u}_t$) for every vintage. Subsequently, `forecast_okun_ssm()` uses
these specific vintage outputs to compute the projected unemployment
path $\hat{u}_{t+1:t+h}$.

## 3. Real-Time Output Gap Estimation

The estimation of the cyclical component requires careful handling of
the HP-filter endpoint bias. While the formal evaluation period begins
in **2015 Q1**, the HP filter is initialized in **2010 Q1**. This
five-year "burn-in" ensures that the GDP trend $\bar{y}_t$ is stabilized
by historical data before the first forecast origin. For parameter
estimation at any time $t$, the filter is applied only to the log-GDP
series $y_{2010:t}$ to ensure no future information after
$\mathcal{I}_t$ enters through the HP filter into the Output gap
$\tilde{y}_t$ into the $\beta$ coefficients. Furthermore, the burn-in
period ensures that the lag terms $\tilde{y}_{t-1}$ and
$\tilde{y}_{t-2}$ are fully populated starting from the first estimation
date.

## 4. Forecasting the Path

The forecasting phase utilizes a different information set for the GDP
gap by concatenating historical log-GDP ($T=2010$ to $t$) with exogenous
GDP forecasts ($T=t+1$ to $t+h$). The HP filter is then applied to this
combined series, and the resulting cyclical components for the $h$
future periods serve as the exogenous inputs for the Okun equation. The
forecast path is generated as:

$$\hat{u}_{t+k} = \rho_t + \beta_{1,t} \cdot \tilde{y}_{t+k} + \beta_{2,t} \cdot \tilde{y}_{t+k-1} + \beta_{3,t} \cdot \tilde{y}_{t+k-2}$$

where $\rho_t$ is the natural rate at the vantage point, $\beta_t$ are
the vintage-specific coefficients, and $\tilde{y}$ is the projected GDP
gap.

## 5. The Evaluation Matrix

The results are stored in a $(\text{Dates} + h) \times \text{Dates}$
matrix, where dates are the dates from $y_t$, the initial vantage point
at the first estimation to $T_{end}$. The columns represent the forecast
origin $t$, while the rows represent the target date $t+k$. On the
Diagonal (where the row index equals the column index), the matrix
contains the actual observed unemployment data. This structure provides
a baseline for visual and statistical error analysis, as the
pseudo-forecasts occupy the cells below the diagonal.
