---
title: "SSM_Documentation"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library("here") # if error, install it
source(here("R", "load_snb_data.R"))
source(here("R", "load_kof_data.R"))
source(here("R", "utils.R"))
source(here("R", "kalman_builder.R"))
source(here("R", "kalman_procedures.R"))
source(here("R", "ssm_forecasting.R"))

```

# SSM Documentation

This Document exists to show how the Kalman filter and State space
models work.

We work with the following notation / specification

**Measurement Equation**
$$y_t = \mu_t + G\rho_t + M\varepsilon_t \tag{1}$$

\vspace{1em}

**Transition / State Equation**
$$\rho_t = \nu_t + H\rho_{t-1} + N\xi_t \tag{2}$$

-   $\mu_t$ is a vector representing the intercept. It can also take on
    exogenous variables
-   $G$ are the factor loadings of the state and relates state and
    measurement equation
-   $M$ is the error var/covar matrix of the measurement equation
-   $\nu_t$ is the intercept of the state euqation, can be used to
    represent a drift
-   $H$ is the transition matrix between the state at $t$ and $t-1$
-   $N$ is the error matrix of the state equation

## Estimation Pipeline

The model uses a coded function of the Kalman Filter available in
`kalmann_procedures.R` This function takes in the matrices that make up
the filter listed above, the data for Y, initial guesses, incic_pos for
when a variable is constrained to be positive, reconcilliationf for
functions / R and Q functions for a quadratic filter which is currently
not used.

The way the matrices are build is that you do that outside of the
filter. You build them including the specified parameters and potential
exogenous data and then pass them into the filter to get the filtered
values and the log likelyhood.

```{r}
# uncomment to see the function
# kalman_filter
```

## Optimization approach

The optimization works with a numeric optimizer from the `optimx`
package. It optimizes an objective functions by passing specific
parameters to it. In the context of the Kalman filter here, the
objective function is a function that sets up the matrices to be passed
into the Kalman filter above, which returns the log likelihood which is
the object that is optimized.

In more detail the flow of the pipeline is: - Determine parameters
(either from first guess, or new guess by optimizer) - Create Kalman
filter matrices from the parameters - Pass the Matrices into the filter,
return likelihood. - optimizer chooses new parameters in response to the
results - when the optimizer doesn't find a good improvement in
parameters, returns them as final estimatiion

```{r}
loglik_ssm
```

The Wrapper function tidies up this loop. You can specify how many loops
you want to run and how many methods you want. Each loop then runs the
number of sleected methods. If you select two it takes the initial
guesses runs the first method and then takes the final output of the
parameters and uses them as the new starting guess for the second method
and so on. Once all methods are done, it goes to the second loop
iteration, still taking the last optimized guess. with this function you
can estimate the parameters for the kalman filter.

```{r}
ssm_optimizer_wrapper
```

# Matrix Handling

The challenge here is how to handle the matrices. Because they need to
be updated with the parameters durnin the estimation, they cannot
directly be constructed outside of the estimation loop. The solution to
this is that we use a function that creates the matrix which takes the
parameters as input and adds creates the full matrix to pass into the
filter. To solve this we use function factories.

Function factories are functions that create functions. The idea here is
that we have have a factory that creates a certain function that already
has the specifications of how many parameters are needed. Therefore we
can fore example specify if there is an AR(1) or a Random walk process.
We can also specify for example how many exogenous variables exist and
their corresponding parameters. The output function of the factory, the
builder function then just takes the parameters from the optimizer and
constructs the final matrix with the parameters from the optimizer

## Buildiung mu_t

This is the most difficult matrix to build, as it can contain exogenous
data. In context of the Kalman filter it is similiar to an intercept. If
it is specified as such it can contain the beta0 as the intercept. if
there is any additional exogenous data it is in mu_t as well.

The final mu_t is a vector containting as many rows as there are
measurement variables Y. The builder returns a matrix / dataframe with
as many columns as observations in t.

The filter then picks the given t at each point.

it looks like this

\$\$ \mu\_t =

\begin{bmatrix}
        
        \beta_0 * 1 + \beta_1 * X_{1,1} + \beta_2 * X_{1,2} && 0 \\
        \beta_0 * 1 + \beta_1 * X_{2,1} + \beta_2 * X_{2,2} && 0 \\
        
        
\end{bmatrix}

\$\$

Where:

$$X\_{1,n}$$ are all observations at time T = 1},$$ X\_{2,n}$$ are all
observations at time T = 2 and so on}

where each row is an observation of time t from 1:T. The filter then
selects the corresponding row at each point in time.

The factoryfunction then returns a function where the parameters (betas)
are replaced by the optimizer guess. Additionally it returns the
manifest which conains the parameters used for an overview of all
parameters that need to be estimated. This list then also contains
educated guesses for the parameters and constraints on the value if they
are needed

```{r}
mu_t_matrix_factory
```

## Building G

The Matrix G governs the link between the state equatin and the
measurement equation. it contains the factory loadings. it's size is
governed by the number of states (nr) and the number of Measurement (Y)
Variables. As default all the values are 1, i.e a 1 to 1 transition.

At the moment there are no parameters gouverning the matrix, so the
parameter input is useless. It exists for a possible future utilization

```{r}
G_matrix_factory
```

## Building M

The matrix $M$ represents the measurement error covariance matrix. It
quantifies the uncertainty or "noise" associated with each observed
variable (Y) in the measurement equation.

An example would be the following

In this case there is independent noise, i.e. only the diagonal is non
0. Further the variance can be non negative, so we have a rule to the
noise taht it is $0< \sigma$. for that we transform any guess of the
optimizer with an exponential funcion `exp()`.

The variances will be estimated. As for interpretation here, if the
variances (chosen by the optimizer) are small, the kalman filter places
a lot of "trust" in the observed variance and the state very directly
tracks them (given the loadings of G). If sigma is high then the
opposite is the case.

```{r}
M_matrix_factory
```

## Buildung H

The H Matrix is the State Transition matrix determining the relationship
between

$\rho_t$ and $\rho_{t-1}$. It's size depends on the number of states.
Currently there is only a specification for 1 state. Currently it can
either be a random walk or an AR(1) process, which has a parameter phi
that has to be estimated.

If the specification is randowm walk it returns

$H = \begin{bmatrix} 1 \end{bmatrix}$

else it returns

$H = \begin{bmatrix} \phi \end{bmatrix}$

The phi has to be $0<\phi < 1$

```{r}
H_matrix_factory
```

The logic of the random walk for the specification in the model is that
the rate is structural and only shocks to the natural rate change it.
These shocks are then governed by the matrix N

## Building N

Currently there is no Function to determine N, it is sepcified directly.
This will be changed in the future. You can set it in the
`kalman_builder.R`. in the `loglik_ssm()`function.

# The Manifest logic

Each of the function returns a manifest which contains the metadata /
the parameter constraints plus an inital guess for each varaible for the
optimizer.

# Rolling Pseudo-Out-of-Sample Forecast Logic

The forecasting framework is implemented in two distinct phases:
**Rolling Parameter Estimation** and **Path Projection**.

## 1. Temporal Framework

We define the time dimension of our sample as $T \in [1, T_{end}]$. For
any given forecast origin $t$, the model parameters are estimated using
the information set $\mathcal{I}_t = \{y_1, \dots, y_t\}$. From this
vantage point (pseudo present where we forecast from), we project the
unemployment rate forward for a horizon $h = \{1, \dots, 8\}$,
corresponding to two years. Once the forecast at origin $t$ is complete,
the vantage point advances to $t+1$, where the model is re-estimated
using $\mathcal{I}_{t+1}$, and a new path is generated. This recursive
process repeats until $\mathcal{I} = T\_{end}$ reaches the end of the
available estimation sample.

## 2. Function Workflow

The logic is split between two primary functions. First,
`get_ssm_forecast_parameters()` executes the rolling estimation,
yielding a sequence of optimal parameter vectors $\theta_t$ and the
filtered state (natural rate) $\rho_t$ (here the natural rate
$\bar{u}_t$) for every vintage. Subsequently, `forecast_okun_ssm()` uses
these specific vintage outputs to compute the projected unemployment
path $\hat{u}_{t+1:t+h}$.

## 3. Real-Time Output Gap Estimation

The estimation of the cyclical component requires careful handling of
the HP-filter endpoint bias. While the formal evaluation period begins
in **2015 Q1**, the HP filter is initialized in **2010 Q1**. This
five-year "burn-in" ensures that the GDP trend $\bar{y}_t$ is stabilized
by historical data before the first forecast origin. For parameter
estimation at any time $t$, the filter is applied only to the log-GDP
series $y_{2010:t}$ to ensure no future information after
$\mathcal{I}_t$ enters through the HP filter into the Output gap
$\tilde{y}_t$ into the $\beta$ coefficients. Furthermore, the burn-in
period ensures that the lag terms $\tilde{y}_{t-1}$ and
$\tilde{y}_{t-2}$ are fully populated starting from the first estimation
date.

## 4. Forecasting the Path

The forecasting phase utilizes a different information set for the GDP
gap by concatenating historical log-GDP ($T=2010$ to $t$) with exogenous
GDP forecasts ($T=t+1$ to $t+h$). The HP filter is then applied to this
combined series, and the resulting cyclical components for the $h$
future periods serve as the exogenous inputs for the Okun equation. The
forecast path is generated as:

$$\hat{u}_{t+k} = \rho_t + \beta_{1,t} \cdot \tilde{y}_{t+k} + \beta_{2,t} \cdot \tilde{y}_{t+k-1} + \beta_{3,t} \cdot \tilde{y}_{t+k-2}$$

where $\rho_t$ is the natural rate at the vantage point, $\beta_t$ are
the vintage-specific coefficients, and $\tilde{y}$ is the projected GDP
gap.

## 5. The Evaluation Matrix

The results are stored in a $(\text{Dates} + h) \times \text{Dates}$
matrix, where dates are the dates from $y_t$, the initial vantage point
at the first estimation to $T_{end}$. The columns represent the forecast
origin $t$, while the rows represent the target date $t+k$. On the
Diagonal (where the row index equals the column index), the matrix
contains the actual observed unemployment data. This structure provides
a baseline for visual and statistical error analysis, as the
pseudo-forecasts occupy the cells below the diagonal.
