---
title: "main"
output: html_document
---


```{r load packages}
#---------i. Clear environment
rm(list=ls())
#---------ii.Load here package to enable finding script loading other package
# MISSING Add Package loader from forecasting prokect
# Maybe using a pip freeze type thing (packagerat?)


# --------- ii. Load here package to enable finding script loading other packages 

# First, define a function that checks if a package is installed. 
# If not, it installs it. Then it loads it.
package_loader <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    message(paste("Installing missing package:", pkg))
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Load here package to be able to locate subsequent files
package_loader("here")

```

```{r}
library(here)
source(here("R", "load_snb_data.R"))
source(here("R", "utils.R"))
```

```{r}
source(here("scripts", "01_load_data.R"))
```

# Clean Data and organzie into one dataframe

# Clean REER

First need to find a way to download it



# Clean CPI Data
```{r}
cpi_ts <- format_time_series_df(cpi_raw, "Datum / Date", "38687", "cpi", "%Y-%m-%d")

```

Decide which series to take (what date is 100)

  select date and values
  rename columns to date and money_market_rate
  turn dates into ym (can always be turned into yq later)
  turn rates into floats(assurance)
  

# Clean Unemployment Data
```{r}
unemployment <- unemployment_raw %>%
  filter(REGION %in% "Total")

unemployment_ts <- format_time_series_df(unemployment, "PERIOD", "VALUE", "unemployment", "%Y-%m")

```


select region total
select date column annd values
rename columns to date and unemp_rate
turn date into ym,
turn values in floats


# Clean employment Data
```{r}





```


```{r}
# 1. The Raw String
quarter_str <- "1991Q3"

# 2. Convert to Year-Quarter object, then to Year-Month
# format = "%YQ%q" tells R that %Y is the year and %q is the quarter number
clean_date <- as.yearmon(as.yearqtr(quarter_str, format = "%YQ%q"))

print(clean_date)
# Output: "Jul 1991"
```



Select 5-96 code (excludes agriculture) and then drop column
select sex total and then drop column
rename columns to date and emp_tot
select dates and values
turn date into yq
turn values into floast 


# clean GDP Data
```{r}
data_gdp <- data_gdp_raw %>% 
  dplyr::filter(
    structure %in% c("gdp","inv_constr","inv_fixed", "cons_priv","cons_gov",          # choose GDP components
                     "exp_good_ex_vm","exp_serv","imp_serv","imp_good_ex_v"),
    type == "real",
    seas_adj == "cssa",
    structure == "gdp"
    
  )  
```

Values already selected in download -> move here and do that in cleaning
still need to:
  drop unnecessary columns: structure, type, seas_adj
  rename columns to date and gdp
  convert date into yq
  convert values in floats





