---
title: "main"
output: html_document
---


```{r load packages}
#---------i. Clear environment
rm(list=ls())
#---------ii.Load here package to enable finding script loading other package
# MISSING Add Package loader from forecasting prokect
# Maybe using a pip freeze type thing (packagerat?)


# --------- ii. Load here package to enable finding script loading other packages 

# First, define a function that checks if a package is installed. 
# If not, it installs it. Then it loads it.
package_loader <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    message(paste("Installing missing package:", pkg))
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Load here package to be able to locate subsequent files
package_loader("here")

```

```{r}
library(here)
source(here("R", "load_snb_data.R"))
source(here("R", "utils.R"))
```

```{r}
source(here("scripts", "01_load_data.R"))
```

# Clean Data and organzie into one dataframe

General questions
  How do I handle dates? Thinking using ym since some data is mothly, handling the missing values with Kalman filter
  But is that good for the compute? Or even reasonable


## Clean SNB Money market data
```{r}
library(tidyverse)

mm_raw_wide <- mm_data %>%
  pivot_wider(
    names_from = "D0",
    values_from = "Value"
  )

saron_ts <- format_time_series_df(mm_raw_wide, "Date", "SARON", "saron", "%Y-%m")
libor_ts <- format_time_series_df(mm_raw_wide, "Date", "3M0", "3m_libor", "%Y-%m")

```

  

# Clean SNB Gov Bond Data
```{r}
gb_raw_wide <- gb_data %>%
  pivot_wider(
    names_from = "D0",
    values_from = "Value"
)


gb_5y_ts <- format_time_series_df(gb_raw_wide, "Date", "5J", "5y_bond", "%Y-%m")
gb_10y_ts <- format_time_series_df(gb_raw_wide, "Date", "10J", "10y_bond", "%Y-%m")

```

Access data from list
  select date and values
  rename columns to date and gov_bond_rate_{duration}y
  turn dates into yq
  turn rates into floats(assurance)
  
  
# Clean REER

First need to find a way to download it



# Clean CPI Data
```{r}
cpi_ts <- format_time_series_df(cpi_raw, "Datum / Date", "38687", "cpi", "%Y-%m-%d")

```

Decide which series to take (what date is 100)

  select date and values
  rename columns to date and money_market_rate
  turn dates into ym (can always be turned into yq later)
  turn rates into floats(assurance)
  

# Clean Unemployment Data
```{r}
unemployment <- unemployment_raw %>%
  filter(REGION %in% "Total")

unemployment_ts <- format_time_series_df(unemployment, "PERIOD", "VALUE", "unemployment", "%Y-%m")

```


select region total
select date column annd values
rename columns to date and unemp_rate
turn date into ym,
turn values in floats


# Clean employment Data
```{r}
emp_raw

employment <- emp_raw %>%
  filter(Division.économique %in% "5-96 Total") %>%
  filter(Taux.d.occupation %in% "Equivalents plein temps, désaisonnalisé") %>%
  filter(Sexe %in% "Sexe - total") 


# Returns NAN Currently pick up here aragin later need to change function to handle quarters or at least return the initial values if the result is NAN
employment_ts <- format_time_series_df(employment,
                                            "Trimestre",
                                            "Emplois.selon.la.division.économique..le.taux.d.occupation.et.le.sexe",
                                            "unemployment",
                                            "%YQ%q")



```

Select 5-96 code (excludes agriculture) and then drop column
select sex total and then drop column
rename columns to date and emp_tot
select dates and values
turn date into yq
turn values into floast 


# clean GDP Data
```{r}

```

Values already selected in download -> move here and do that in cleaning
still need to:
  drop unnecessary columns: structure, type, seas_adj
  rename columns to date and gdp
  convert date into yq
  convert values in floats






